A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-07-01 _manuscript/_tex/manuscript.tex
update _freeze and _manuscript

To our knowledge, existing approaches for sample size planning for GLMMs
have exclusively focused on planning for power. In Table 1, we review
available software packages that can be used to perform power analysis
for multilevel models in R {[}Version 4.3.3; R Core Team
(\citeproc{ref-R-base}{2024}){]}. Power analysis methods can be
categorized into formula-based, summary-statistics-based and
simulation-based methods
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022}). Formula-based methods rely on exact formulas to
calculate power directly. Summary-statistics-based methods use
statistical theory to approximate power based on formula-based methods
developed for simple t-tests. Simulation-based methods rely on
repeatedly simulating data with a known true effect size and estimating
power empirically, that is what percentage of simulated datasets
produces a significant p-value. Available formula-based and
summary-statistics-based software packages for multilevel models often
do not include GLMMs or are limited to simple designs
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022};
\citeproc{ref-westfallStatisticalPowerOptimal2014}{Westfall et al.,
2014}), making it necessary to build data simulations tailored
specifically to the study design. A number of tutorials have been
published describing how to perform such simulation-based power analysis
for multilevel models
(\citeproc{ref-arendStatisticalPowerTwolevel2019}{Arend \& Schäfer,
2019}; \citeproc{ref-brysbaertPowerAnalysisEffect2018}{Brysbaert \&
Stevens, 2018};
\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}; \citeproc{ref-greenSIMRPackagePower2016a}{Green \& MacLeod,
2016a}; \citeproc{ref-johnsonPowerAnalysisGeneralized2015}{Johnson et
al., 2015}; \citeproc{ref-kainPracticalGuidePower2015}{Kain et al.,
2015}; \citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al.,
2021}; \citeproc{ref-lafitSelectionNumberParticipants2021}{Lafit et al.,
2021}; \citeproc{ref-zimmerSampleSizePlanning2023}{Zimmer et al.,
2023}). However, many of these tutorials focus on LMMs and the most
common study designs (see
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}
for a tutorial that also covers more advanced settings). This narrow
focus provides limited guidance for researchers using more complex study
designs, especially when little prior knowledge about plausible effect
sizes is available (see the discussion in
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}).
Simulation-based power analysis with GLMMs requires making a range of
assumptions about the model structure that should align with the
characteristics of the data being analyzed. Existing tutorials often
rely on heuristics for specifying variance components (e.g., the
standard deviation of random intercepts) or assume that results from
meta-analyses or data from pilot studies are available to determine
plausible values for all model parameters. However, in practice,
knowledge about those parameters from prior studies is often limited,
which makes specifying assumptions a practical challenge (see the
discussion in Maxwell et al.
(\citeproc{ref-maxwellSampleSizePlanning2008}{2008}) and Kumle et al.
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{2021})). We will
discuss a number of strategies on how to specify model parameters for
application-specific, tailored data simulations in a later chapter.

@@

\multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{longpower}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.57in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Formula-based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{power}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{analysis;}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{focus}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{on}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{longitudinal}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{data}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{0000FF}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\href{https://atrihub.shinyapps.io/power/}{yes}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{limited}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{set}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{of}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{study}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{designs}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{with}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{two}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{levels}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.79in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Iddi}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\&}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{C}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Donohue}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{(2022)}}}} \\

@@

\multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{pass.lme}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.57in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Formula-based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{power}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{analysis;}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{limited}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{documentation}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{none}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.79in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Marco}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Chak}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Yan}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{YU}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{(2019)}}}} \\

@@

\paragraph{THEORY.}\label{theory-1}

@@

\paragraph{PRACTICE.}\label{practice-1}

@@

\subparagraph{Our specific GLMM.}\label{our-specific-glmm}

In a GLMM, the expected value of the dependent variable \(Y\)
conditioned on the vector of predictor variables \(\mathbf{X}\) and
random effects \(\mathbf{U}\), transformed by a link function \(g()\) is
modeled as a linear combination \(\eta\) of the predictor variables
\(\mathbf{X}\), the random effects \(\mathbf{U}\), and the model
parameters \(\mathbf{\beta}\)
(\citeproc{ref-fahrmeirRegressionModelsMethods2021}{Fahrmeir et al.,
2021}): \[
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
\] Equivalently, the conditional expected value is modeled as the linear
combination \(\eta\), transformed by the inverse link function
\(g^{-1}()\): \[
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u}) = g^{-1}(\eta)
\] If the dependent variable (i.e., diagnostic decision) \(Y\) is a
binary variable with values \(0\) (i.e., inaccurate), or \(1\) (i.e.,
accurate), the conditional expected value is equivalent to the
probability: \[
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
\] In our case study, \(P_{si}\) is the conditional probability that
subject \(s\) gives the correct response to item (i.e., CT scan) \(i\).

@@

\subparagraph{Simulation function in R.}\label{simulation-function-in-r}

The following R function simulates a full dataset structured according
to the design of our case study.

@@

\paragraph{THEORY.}\label{theory-3}

@@

\paragraph{PRACTICE.}\label{practice-3}

@@

\paragraph{THEORY.}\label{theory-4}

@@

\paragraph{PRACTICE.}\label{practice-4}

@@

For the next plot, we have set the item standard deviation to almost
zero (\(\sigma_I = 0.01\)). This gives us a better way to see the
variability between persons. As an example, Figure~\ref{fig-margdist2}
reveals a number of implicit assumptions about the comparison between
experts and students: With incorrect advice, virtually all experts have
a higher probability of making a correct diagnosis compared to students
when considering only items with average difficulty. In contrast, there
is considerable overlap in probability between experts and students with
no advice and even higher overlap with correct advice. Patterns like
these should be considered carefully and discussed with the domain
experts. Parameter values (\(\beta\) parameters, and \(\sigma_S\))
should be adjusted if the implications do not seem reasonable. We could
also have a closer look at the variability between items by setting the
subject standard deviation to almost zero (\(\sigma_S = 0.01\)).

The final plot demonstrates that these plots are also useful for
spotting standard deviations that were specified too high. For
Figure~\ref{fig-margdist3} we have set \(\sigma_S = 3\) and
\(\sigma_I = 3\). This implies that in each experimental condition, the
probabilities that a subject solves an item are overwhelmingly close to
either 0 or 1 and nothing in between, which is not a plausible
assumption. These high standard deviations do not account for the
inherent variability and complexity of human performance. For example,
the expectation that an expert with low ability and incorrect advice
would solve a difficult item with a probability close to zero is not
convincing.

@@

\paragraph{THEORY.}\label{theory-5}

@@

\subparagraph{PRACTICE.}\label{practice-5}

@@

\paragraph{Hypothesis testing.}\label{hypothesis-testing}

If we chose hypothesis testing for our case study, we would test a
combined null hypothesis \(H_0\) that consists of four separate null
hypotheses: \[
\begin{aligned}
H_{01}:\ & P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{02}:\ &P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{03}:\ &P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \\
H_{04}:\ & P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0)
\end{aligned}
\] The combined null hypothesis \(H_0\) should only be rejected if
\textbf{all} individual null hypotheses are rejected (i.e.,
intersection-union setting;
\citeproc{ref-dmitrienkoTraditionalMultiplicityAdjustment2013}{Dmitrienko
\& D'Agostino, 2013}). In such cases, the error probabilities do not
accumulate, and we would waste power when correcting for multiple tests.

@@

\paragraph{Interval estimation.}\label{interval-estimation}

If we chose parameter estimation for our case study, we would focus on
the two-sided CIs of the four contrasts of interest. With a standard
confidence level of \(1 - \alpha = 0.95\) plausible values are clearly
in the positive range for the first, third, and fourth contrast, while
both negative and positive values seem plausible for the second
contrast. Note that due to the constrained range of the probability
scale, the width of the CI differs between the four contrasts (which is
the expected behavior for binomial GLMMs). The smallest width is
observed for the first contrast (expert with correct advice vs.~expert
without advice), where both underlying probabilities are close to 1. The
largest width is observed for the fourth contrast (student with
incorrect advice vs.~student without advice), where both underlying
probabilities are closer to 0.5.

@@

\subsection{Step 7 (optional): Sensitivity
analysis}\label{step-7-optional-sensitivity-analysis}

>>>>

To our knowledge, existing approaches for sample size planning for GLMMs
have exclusively focused on planning for power. In Table 1, we review
available software packages that can be used to perform power analysis
for multilevel models in R {[}Version 4.3.3; R Core Team
(\citeproc{ref-R-base}{2024}){]}. Power analysis methods can be
categorized into formula-based, summary-statistics-based and
simulation-based methods
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022}). Formula-based methods rely on exact formulas to
calculate power directly. Summary-statistics-based methods use
statistical theory to approximate power based on formula-based methods
developed for simple t-tests. Simulation-based methods rely on
repeatedly simulating data with a known true effect size and estimating
power empirically, that is what percentage of simulated datasets
produces a significant p-value. Available formula-based and
summary-statistics-based software packages for multilevel models often
do not include GLMMs or are limited to simple designs
(\citeproc{ref-murayamaSummarystatisticsbasedPowerAnalysis2022}{Murayama
et al., 2022};
\citeproc{ref-westfallStatisticalPowerOptimal2014}{Westfall et al.,
2014}), making it necessary to build data simulations tailored
specifically to the study design. A number of tutorials have been
published describing how to perform such simulation-based power analysis
for multilevel models
(\citeproc{ref-arendStatisticalPowerTwolevel2019}{Arend \& Schäfer,
2019}; \citeproc{ref-brysbaertPowerAnalysisEffect2018}{Brysbaert \&
Stevens, 2018};
\citeproc{ref-debruineUnderstandingMixedEffectsModels2021}{DeBruine \&
Barr, 2021}; \citeproc{ref-greenSIMRPackagePower2016}{Green \& MacLeod,
2016}; \citeproc{ref-johnsonPowerAnalysisGeneralized2015}{Johnson et
al., 2015}; \citeproc{ref-kainPracticalGuidePower2015}{Kain et al.,
2015}; \citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al.,
2021}; \citeproc{ref-lafitSelectionNumberParticipants2021}{Lafit et al.,
2021}; \citeproc{ref-zimmerSampleSizePlanning2023}{Zimmer et al.,
2023}). However, many of these tutorials focus on LMMs and the most
common study designs (see
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}
for a tutorial that also covers more advanced settings). This narrow
focus provides limited guidance for researchers using more complex study
designs, especially when little prior knowledge about plausible effect
sizes is available (see the discussion in
\citeproc{ref-kumleEstimatingPowerGeneralized2021}{Kumle et al., 2021}).
Simulation-based power analysis with GLMMs requires making a range of
assumptions about the model structure that should align with the
characteristics of the data being analyzed. Existing tutorials often
rely on heuristics for specifying variance components (e.g., the
standard deviation of random intercepts) or assume that results from
meta-analyses or data from pilot studies are available to determine
plausible values for all model parameters. However, in practice,
knowledge about those parameters from prior studies is often limited,
which makes specifying assumptions a practical challenge (see the
discussion in Maxwell et al.
(\citeproc{ref-maxwellSampleSizePlanning2008}{2008}) and Kumle et al.
(\citeproc{ref-kumleEstimatingPowerGeneralized2021}{2021})). We will
discuss a number of strategies on how to specify model parameters for
application-specific, tailored data simulations in a later chapter.

@@

\multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{longpower}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.57in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Formula-based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{power}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{analysis;}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{focus}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{on}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{longitudinal}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{data}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{0000FF}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\href{https://atrihub.shinyapps.io/power/}{yes}}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{limited}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{set}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{of}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{study}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{designs}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{with}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{two}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{levels}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.79in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Iddi}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\&}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Donohue}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{(2022)}}}} \\

@@

\multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{pass.lme}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.57in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Formula-based}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{power}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{analysis;}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{limited}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{documentation}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.59in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{no}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 1.18in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{none}}}} & \multicolumn{1}{>{\centering}m{\dimexpr 0.79in+0\tabcolsep}}{\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{Yu}}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{\ }}}\textcolor[HTML]{000000}{\fontsize{10}{20}\selectfont{\global\setmainfont{Times New Roman}{(2019)}}}} \\

@@

\subsubsection{THEORY}\label{theory-1}

@@

\subsubsection{PRACTICE}\label{practice-1}

@@

\textbf{Our specific GLMM.} In a GLMM, the expected value of the
dependent variable \(Y\) conditioned on the vector of predictor
variables \(\mathbf{X}\) and random effects \(\mathbf{U}\), transformed
by a link function \(g()\) is modeled as a linear combination \(\eta\)
of the predictor variables \(\mathbf{X}\), the random effects
\(\mathbf{U}\), and the model parameters \(\mathbf{\beta}\)
(\citeproc{ref-fahrmeirRegressionModelsMethods2021}{Fahrmeir et al.,
2021}): \[
g(E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})) = \eta
\] Equivalently, the conditional expected value is modeled as the linear
combination \(\eta\), transformed by the inverse link function
\(g^{-1}()\): \[
E(Y|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u}) = g^{-1}(\eta)
\] If the dependent variable (i.e., diagnostic decision) \(Y\) is a
binary variable with values \(0\) (i.e., inaccurate), or \(1\) (i.e.,
accurate), the conditional expected value is equivalent to the
probability: \[
P_{si} := P(Y = 1|\mathbf{X}=\mathbf{x},\mathbf{U}=\mathbf{u})
\] In our case study, \(P_{si}\) is the conditional probability that
subject \(s\) gives the correct response to item (i.e., CT scan) \(i\).

@@

\textbf{Simulation function in R.} The following R function simulates a
full dataset structured according to the design of our case study.

@@

\subsubsection{THEORY}\label{theory-3}

@@

\subsubsection{PRACTICE}\label{practice-3}

@@

\subsubsection{THEORY}\label{theory-4}

@@

\subsubsection{PRACTICE}\label{practice-4}

@@

For the next plot, we have set the item standard deviation to almost
zero (\(\sigma_I = 0.01\)). This gives us a better way to see the
variability between persons. As an example, Figure~\ref{fig-margdist2}
reveals a number of implicit assumptions about the comparison between
experts and students: With incorrect advice, virtually all experts have
a higher probability of making a correct diagnosis compared to students
when considering only items with average difficulty. In contrast, there
is considerable overlap in probability between experts and students with
no advice and even higher overlap with correct advice. Patterns like
these should be considered carefully and discussed with the domain
experts. Parameter values (\(\beta\) parameters, and \(\sigma_S\))
should be adjusted if the implications do not seem reasonable.\\
The final plot demonstrates that these plots are also useful for
spotting standard deviations that were specified too high. For
Figure~\ref{fig-margdist3} we have set \(\sigma_S = 3\) and
\(\sigma_I = 3\). This implies that in each experimental condition, the
probabilities that a subject solves an item are overwhelmingly close to
either 0 or 1 and nothing in between, which is not a plausible
assumption. These high standard deviations do not account for the
inherent variability and complexity of human performance. For example,
the expectation that an expert with low ability and incorrect advice
would solve a difficult item with a probability close to zero is not
convincing.

@@

\subsubsection{THEORY}\label{theory-5}

@@

\subsubsection{PRACTICE}\label{practice-5}

@@

\textbf{Hypothesis testing.} If we chose hypothesis testing for our case
study, we would test a combined null hypothesis \(H_0\) that consists of
four separate null hypotheses: \[
\begin{aligned}
H_{01}:\ & P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{02}:\ &P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 1, u_{0s} = 0, u_{0i} = 0) \\
H_{03}:\ &P(Y=1|advice\_present = 1, advice\_correct = 1, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \\
H_{04}:\ & P(Y=1|advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) \leq \\
& P(Y=1|advice\_present = 1, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0)
\end{aligned}
\] The combined null hypothesis \(H_0\) should only be rejected if
\textbf{all} individual null hypotheses are rejected (i.e.,
intersection-union setting;
\citeproc{ref-dmitrienkoTraditionalMultiplicityAdjustment2013}{Dmitrienko
\& D'Agostino, 2013}). In such cases, the error probabilities do not
accumulate, and we would waste power when correcting for multiple tests.

@@

\textbf{Interval estimation.} If we chose parameter estimation for our
case study, we would focus on the two-sided CIs of the four contrasts of
interest. With a standard confidence level of \(1 - \alpha = 0.95\)
plausible values are clearly in the positive range for the first, third,
and fourth contrast, while both negative and positive values seem
plausible for the second contrast. Note that due to the constrained
range of the probability scale, the width of the CI differs between the
four contrasts (which is the expected behavior for binomial GLMMs). The
smallest width is observed for the first contrast (expert with correct
advice vs.~expert without advice), where both underlying probabilities
are close to 1. The largest width is observed for the fourth contrast
(student with incorrect advice vs.~student without advice), where both
underlying probabilities are closer to 0.5.

@@

\subsection{Step 7 (optional): Conduct sensitivity
analysis}\label{step-7-optional-conduct-sensitivity-analysis}
